Quantifying the Importance of Change for Understanding Differences in Personal Culture

### 2024-03-22 manuscript/manuscript.tex

An important contemporary debate in sociology centers on the importance
of intrapersonal change for understanding differences in ``personal
culture'' -- attitudes, beliefs, values, and practices (Kiley and Vaisey
2020; Lersch 2023; Lizardo 2017; Vaisey and Lizardo 2016;
\textbf{underwood2021?}). This question has deep roots in seemingly
contradictory theoretical perspectives. For example, pragmatist theories
of action claim that changes in social environments cause people to
adapt their views and make new cultural meanings (Gross 2009; Swidler
2001), while Bourdieusian practice theories argue that the ``past
conditions of production'' leave a mark on people's personal culture
that lasts throughout their lives (Bourdieu 1990). Models of social
influence assume that people adapt their culture in the face of new
information (Goldberg and Stein 2018), while the emphasis on cohort
effects in models of aggregate social and cultural change requires them
to be open to change while young but become fairly resistant to it as
they age (Ryder 1965). Finally, life course theories posit important
changes over time as people advance through important transitions in
their lives, but that certain life stages have more lasting influence
than others (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Recent work on this topic has converged on a general conclusion that
people do sometimes make substantial shifts in their personal culture
over time but that most elements of personal culture are characterized
by low rates of persistent change during adulthood (Kiley and Vaisey
2020; Lersch 2023). Until recently, this work has adopted a ``tournament
of models'' approach to adjudicating different theoretical perspectives
(Lersch 2023: p.~228). In this framework, researchers formalized
different models of change over time and relied on model selection
criteria---primarily the Bayesian Information Criterion (Raftery
1995)---to classify different questions as demonstrating either ``no
change'' or ``any change.'' The size of these piles then serves as the
primary evidence for the truth of different arguments (Kiley and Vaisey
2020; Vaisey and Kiley 2021).

@@

In a recent paper, Lersch (2023) takes an important step forward by
devising a Life Course Adaption Model (LCAM), a mixed-effects growth
curve model that quantifies various components of cultural difference:
early imprinting, persistent change, biographical experience, and
fluctuation. In quantifying these aspects of cultural difference, the
LCAM lets researchers address a broader range of questions than the
single question of whether a measure is characterized by stability or
change. However, because the LCAM produces so many different measures,
it is not obvious how to use these components to understand the relative
importance of change, espeically when change is measured in non-linear
ways.

In this paper, we build on the LCAM by outlining a single metric that
combines diverse components of the LCAM to quantify the relative
importance of intrapersonal change and interpersonal difference for
explaining variation in a population for a variable over time. Our
approach isolates the ``systematic'' components of variance (i.e., it
removes random change and measurement error), to provide a single
succinct summary of how much systematic variation can reasonably be
attributed to people making persistent changes over time. We demonstrate
the utility of this quantity in two ways.

@@

\hypertarget{expectations}{%
\subsection{Expectations}\label{expectations}}

By shifting the question and adjusting the approach, we expect
consistent patterns across datasets. First, because previous work found
evidence of durable change on most measures of personal culture, we
expect that we will find some intrapersonal change in most items. But
because these earlier studies also found many questions (between 25 and
35 percent) lacking evidence of \emph{any} durable change, we expect
considerable variation in the proportion of variance accounted for by
intrapersonal change, with some questions indicating essentially none at
all.

Second, we expect the variance attributable to intrapersonal change to
be modest compared to pre-existing interpersonal differences. This
expectation should be uncontroversial, regardless of the theoretical
process assumed to be most relevant for a particular question. Panel
surveys do not encompass individuals' entire lifetimes. Thus, even if
intrapersonal change during adulthood is the major driver of cultural
differences, the change detected in a panel covering 20 years probably
will not fully account for the entire variance of interpersonal
differences.

Our expectations so far largely echo what was observed in previous
studies. However, theoretical perspectives differ in how much change
they expect, enabling quantification to provide a new perspective on
existing questions. ``Formative period'' and cohort theories, which
emphasize early-life socialization as a strong influence on adult
personal culture, imply low amounts of intrapersonal change, even when
we account for the duration of observation. Conversely, other
theoretical perspectives emphasizing the importance of people's
contemporaneous social environment might expect a larger degree of
intrapersonal change, even in a relatively short panel.

Third, we expect the variance explained by each process---interpersonal
differences and intrapersonal change---to depend on specific attributes
of the question and panel. First, the higher the response resolution for
a given question and the larger the sample, the greater the variance
attributed to intrapersonal change should be, as smaller changes in
intrapersonal change will be easier to detect. Second, life course
theories posit that social transitions drive changes in personal
culture. Since the probability that any person experiences such a
transition goes up the longer we observe them, we expect panel duration
to be positively related to the proportion of variation explained by
intrapersonal change.

\hypertarget{analytic-strategy}{%
\section{Analytic Strategy}\label{analytic-strategy}}

@@

\hypertarget{analysis-steps}{%
\subsection{Analysis Steps}\label{analysis-steps}}

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then derive \(\omega\)
for each of these items.

As a second step in our analysis, we use linear regression to model
\(\omega\) as a function of features of the question and panel. This
allows us to explore how these features are associated with the degree
of intrapersonal change. These covariates include the number of response
options (measured as 2, 3-5, and more than 5), the specific survey, the
log of the number of participants, the number of waves observed, the
time period (i.e., decade) the question was first asked, and the total
duration of time the question was asked.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{variance-decomposition}{%
\subsection{Variance Decomposition}\label{variance-decomposition}}

@@

\hypertarget{meta-analysis}{%
\subsection{Meta-Analysis}\label{meta-analysis}}

@@

\hypertarget{college-and-change-in-political-culture}{%
\subsection{College and Change in Political
Culture}\label{college-and-change-in-political-culture}}

These results mostly re-frame and align previous findings, but the value
of our approach lies in its ability to extend the debate to a broader
set of theoretical questions. To show this potential, we turn to a
specific empirical example: the relative importance of intrapersonal
change and interpersonal difference for explaining cultural variation by
level of education.

@@

There are theoretical reasons to believe that education might be
associated with either more or less intrapersonal change. On one hand,
because college graduates are more connected to mainstream discourse and
elite signals, they might be more likely to make durable changes in
response to the emergence of new information, new issues, or political
realignments, while those without chronic information might display more
variance around an unchanging baseline (Zaller 1992). Conversely, it
could be that the observed stability of the college educated reflects
the fact that they have already formed durable opinions and are
relatively closed off to new information. If this is true, college could
be understood as a formative experience that solidifies some dimensions
of personal culture. Perhaps for those who do not attend college, later
life experiences might prove more important in forming or changing
personal culture, as these experiences potentially provide information
that college-educated peers have already received.

@@

@@

On all eight questions presented in Figure 4, \(\omega\) values are
smaller for people with a college degree, meaning intrapersonal change
is less common among them. There are also large differences in
\(\omega\) values across questions for both groups. For example, on the
question of whether the government should try to solve more problems or
leave those problems to be solved by private businesses (``government do
more or less''), less than 1 percent of the systematic variance is
attributable to intrapersonal change for both groups. In other words,
while people might vacillate on this question at random (37 percent of
variance is residual for this question), there is functionally no
evidence that people make systematic changes of opinion on this issue
during the GSS panel.

In contrast, partisan identification and political ideology both show
larger values of \(\omega\) than most other questions, as well as a
larger absolute difference by degree status. Compared to the other
questions, intrapersonal change plays a much larger role in accounting
for partisan identification and political ideology. And this is
particularly true for respondents who do not have a college degree; the
\(\omega\) value for ideological identification among non-college
educated respondents is almost four times that of college-educated
respondents.

@@

The primary objective of this paper is to introduce a new quantity that
can advance debate on whether people change their personal culture --
their attitudes, beliefs, values, and practices -- as they move through
their adult life. Instead of falling back on a ``tournament of models''
approach (Lersch 2023: p.~228) to proclaim victory for a particular
answer to this question, our goal is to refocus the debate to the
theoretically more productive question of the relative importance of
intrapersonal change for explaining differences in personal culture. We
propose a new approach that quantifies the amount of observed systematic
variance that is attributable to either interpersonal differences at
baseline or intrapersonal change over the duration of a panel.

Applying our proposed measure, \(\omega\), to 609 survey items from all
of the panel datasets previously discussed in this debate revealed a
consistent pattern. Nearly all questions show evidence of people making
some durable, intrapersonal change over time. Some show notably high
amounts of intrapersonal change. For some questions about life
satisfaction or views on government benefits, it seems plausible that
differences in adult experiences predominantly account for observed
differences between people.

However, intrapersonal change is often substantially less pronounced
than interpersonal differences, accounting for less than 10 percent of
systematic variance on average across all datasets. On numerous
questions, like those on civil liberties, abortion, generalized trust,
and civic duty, the systematic variance attributable to intrapersonal
change is essentially zero. For these questions, it seems there is not
enough cultural change during adulthood to warrant attributing the
differences we observe to experiences and social transitions; instead,
the primary source of the observed differences appears to stem from
experiences in childhood, adolescence, and early adulthood.

More than adjudicating past positions in this debate, the approach we
presented in this paper will enable researchers to explore more
precisely the relative contributions of change and stability in
explaining variation in personal culture, as well as how these
contributions might differ across populations. To illustrate this, we
showed how the amount of cultural difference explained by intrapersonal
change varies substantially by survey item but also by individuals'
characteristics such as their education. While age and other factors
related to college completion might confound this pattern, it suggests
that college completion may crystallize personal culture to an extent
that renders later adult experiences on political dispositions less
influential.

@@

Despite these limitations, we believe our method and findings have
important implications for social science research. Sociologists
interested in understanding cultural differences have largely asked
about the \emph{existence} of cultural change in adults (e.g., Kiley and
Vaisey 2020; Lersch 2023; Vaisey and Kiley 2021). In any population,
\emph{some} degree of adult cultural change is inevitable. Although
ideal types like the settled dispositions and active updating models are
useful, no theoretical perspective would expect either early life
socialization or adult intrapersonal change to be the sole source of
one's personal culture. Our results reinforce this point, showing the
relevance of both factors and allowing their precise quantification.

This and other recent findings (Quinn et al. 2023; Stewart and Berkman
2023; \textbf{quinn2024?}) suggest that it is theoretically more
productive to measure the relative importance of these two components in
concrete, substantive cases. Drawing a unified conclusion from survey
items measuring various cultural forms on different scales and across
different time frames is challenging. Nevertheless, the general pattern
suggests that, for most items, intrapersonal change in adulthood is not
the primary reason for the differences we see between people in the
world.

We are not claiming that adults remain static or that their changes are
inconsequential. Even quantitatively minor shifts, such as a 2
percentage-point change in support for gay marriage, can have massive
ramifications. While the majority may remain consistent in their views,
understanding the underlying mechanisms of even such numerically minor
shifts remains a crucial task for the sociology of culture.

@@

Our findings also have implications for understanding cultural change at
the aggregate level. Given that intrapersonal change accounts for a
relatively small amount of variance among adults, many forms of cultural
change at the aggregate level are necessarily more likely driven cohort
replacement than by contemporaneous social conditions (Underwoood et al.
2022; Vaisey and Lizardo 2016). This likely holds true for cultural
change at the macro societal level as well as the micro level such as
within organizations, political parties, and professions. Our findings
also indicate that the impact of formative experiences, as opposed to
contemporaneous social factors, on cultural change is contingent upon an
individuals' level of education. The weight of these two factors might
thus change as the prominence of education shifts in the life course. If
scholars find similar differences across other social categorizations
like class or race, it might necessitate a more thorough integration of
demographic processes into the analysis of cultural change at the
population level.

@@

@@

>>>>

An important contemporary debate in sociology centers on the importance
of individual change over time for understanding differences in
``personal culture''---attitudes, beliefs, values, and practices (Kiley
and Vaisey 2020; Lersch 2023; Lizardo 2017; Underwoood et al. 2022;
Vaisey and Lizardo 2016). This question has deep roots in seemingly
contradictory theoretical perspectives. For example, pragmatist theories
of action claim that changes in social environments cause people to
adapt their views and make new cultural meanings (Gross 2009; Swidler
2001), while Bourdieusian practice theories argue that the ``past
conditions of production'' leave a mark on people's personal culture
that lasts throughout their lives (Bourdieu 1990). Models of social
influence assume that people adapt their culture in the face of new
information (Goldberg and Stein 2018), while the emphasis on cohort
effects in models of aggregate social and cultural change requires them
to be open to change while young but become fairly resistant to it as
they age (Ryder 1965). Finally, life course theories posit important
changes over time as people advance through important transitions in
their lives, but that certain life stages have more lasting influence
than others (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Recent work on this topic has converged on a general conclusion that
people do sometimes make substantial shifts in their personal culture
over time but that most elements of personal culture are characterized
by low rates of persistent change during adulthood (Kiley and Vaisey
2020; Lersch 2023). Work leading to this conclusion adopted a
``tournament of models'' approach to adjudicating different theoretical
perspectives (Lersch 2023: p.~228). In this framework, researchers
formalized different models of change over time and relied on model
selection criteria---primarily the Bayesian Information Criterion
(Raftery 1995)---to classify different questions as demonstrating either
``no change'' or ``any change.'' The size of these piles then serves as
the primary evidence for the truth of different arguments (Kiley and
Vaisey 2020; Vaisey and Kiley 2021).

@@

In a recent paper, Lersch (2023) takes an important step forward by
devising a Life Course Adaption Model (LCAM), a mixed-effects growth
curve model that quantifies various components of cultural difference:
early imprinting, persistent change, biographical experience, and
fluctuation. In quantifying these aspects of cultural difference, the
LCAM lets researchers address a broader range of questions than the
single question of whether a measure is characterized by stability or
change. However, because the LCAM produces so many different measures,
it is not obvious how to use these components to understand the relative
importance of within-person change, especially when change is measured
in non-linear ways.

In this paper, we build on the LCAM by outlining a single metric that
combines components of the LCAM to quantify the relative importance of
intrapersonal change and interpersonal difference for explaining
variation in a population for a variable over time. Our approach
isolates the ``systematic'' components of variance (i.e., it removes
random change and measurement error), to provide a single succinct
summary of how much systematic variation can reasonably be attributed to
people making persistent changes over time. This quantity can then be
used as a quick diagnostic to compare across questions, groups, periods,
or samples to isolate where within-person change is more or less useful
for explaining variation. We demonstrate the utility of this quantity in
two ways.

@@

\hypertarget{example-1-cross-question-comparisons}{%
\section{Example 1: Cross-Question
Comparisons}\label{example-1-cross-question-comparisons}}

To demonstrate the utility of this metric, we provide two examples. In
the first, we compare the proportion of systematic variance explained by
intrapersonal change across a broad array of questions.

@@

\hypertarget{analysis}{%
\subsection{Analysis}\label{analysis}}

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then derive \(\omega\)
for each of these items. As a second step, we use linear regression to
model \(\omega\) as a function of features of the question and panel.
This allows us to explore how these features are associated with the
degree of intrapersonal change. These covariates include the number of
response options (measured as 2, 3-5, and more than 5), the specific
survey, the log of the number of participants, the number of waves
observed, the time period (i.e., decade) the question was first asked,
and the total duration of time the question was asked.

\hypertarget{results}{%
\subsection{Results}\label{results}}

@@

@@

\hypertarget{example-2-college-and-change-in-political-culture}{%
\section{Example 2: College and Change in Political
Culture}\label{example-2-college-and-change-in-political-culture}}

These results mostly re-frame and align previous findings, but the value
of our approach lies in its ability to extend the debate to a broader
set of theoretical questions. As a second empirical application of
\(\omega\), we explore the relative importance of intrapersonal change
and interpersonal difference for explaining cultural variation by level
of education.

@@

There are theoretical reasons to believe that education might be
associated with either more or less intrapersonal change over time. On
one hand, because college graduates are more connected to mainstream
discourse and elite signals, they might be more likely to make durable
changes in response to the emergence of new information, new issues, or
political realignments, while those without chronic information might
display more variance around an unchanging baseline (Zaller 1992).
Conversely, because people with less education appear more likely to
engage in automatic cognition in social situations (including survey
interviews) than people with more education (Brett and Miles 2021), they
might be more influenced by changes in the social environment, producing
more temporary change and more durable change. It could be that the
observed stability of the college educated also reflects the fact that
they have already formed durable opinions across various issues and are
relatively closed off to new information. If this is true, college could
be understood as a formative experience that solidifies some dimensions
of personal culture. Perhaps for those who do not attend college, later
life experiences might prove more important in forming or changing
personal culture, as these experiences potentially provide information
that college-educated peers have already received.

\hypertarget{data-1}{%
\subsection{Data}\label{data-1}}

@@

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

@@

On all eight questions presented in Figure 4, \(\omega\) values are
smaller for people with a college degree, meaning intrapersonal change
is less important for explaining variation among them. There are also
large differences in \(\omega\) values across questions for both groups.
For example, on the question of whether the government should try to
solve more problems or leave those problems to be solved by private
businesses (``government do more or less''), less than 1 percent of the
systematic variance is attributable to intrapersonal change for both
groups. In other words, while people might vacillate on this question at
random (37 percent of variance is residual for this question), there is
functionally no evidence that people make systematic changes of opinion
on this issue during the GSS panel.

In contrast, partisan identification and political ideology both show
larger values of \(\omega\) than most other questions, as well as a
larger absolute difference by degree status. Compared to the other
questions, intrapersonal change plays a much larger role in accounting
for variation in partisan identification and political ideology. And
this is particularly true for respondents who do not have a college
degree; the \(\omega\) value for ideological identification among
non-college educated respondents is almost four times that of
college-educated respondents.

@@

The primary objective of this paper is to introduce a new quantity that
can advance debate on whether people change their personal culture --
their attitudes, beliefs, values, and practices -- as they move through
their adult life. Instead of falling back on a ``tournament of models''
approach (Lersch 2023: p.~228) to proclaim a yes or no answer to this
question, or only quantifying individual components of cultural
difference to show how much people change, our metric contextualizes the
amount of observed systematic variance that is attributable to
intrapersonal change over the duration of a panel. In doing so, it
provides a tool to identify questions, groups, or periods where change
is a more meaningful explanation of cultural difference.

We demonstrated the utility of this measure in two ways. First, applying
our proposed measure, \(\omega\), to 609 survey items from the panel
datasets previously studied in this debate revealed a consistent
pattern. Nearly all questions show evidence of people making some
durable, intrapersonal change over time. Some show notably high amounts
of intrapersonal change. For some questions about life satisfaction or
views on government benefits, it seems plausible that differences in
adult experiences predominantly account for observed differences between
people. However, intrapersonal change is often substantially less
pronounced than interpersonal differences, accounting for less than 10
percent of systematic variance on average across all datasets. On
questions related to civil liberties, abortion, generalized trust, and
civic duty, systematic variance attributable to intrapersonal change is
essentially zero. For these questions, it seems there is not enough
cultural change during adulthood to warrant attributing the differences
we observe to experiences and social transitions; instead, the primary
source of the observed differences appears to stem from experiences in
childhood, adolescence, and early adulthood.

While these findings generally echo the general consensus that has
emerged from previous work, it adds two dimensions. First, it quantifies
the relative importance of change across these questions, providing more
resolution within what were mainly piles of questions either showing
evidence of change or no evidence of change. This allows for greater
comparison across questions and it provides a benchmark for researchers
exploring new questions. Second, our metric can help contextualize the
quantities identified in the LCAM. Is .07 standard deviations of change
every 10 years a lot of change? For some questions, our metric will say
this change explains a large component of variance, and for others it
will not.

In a second application of the quantity, we showed how it can open up
new lines of inquiry. We showed how the amount of cultural difference
explained by intrapersonal change varies substantially not just by
survey item but also by individuals' characteristics such as their
education. While age and other factors related to college completion
might confound this pattern, it suggests that college completion may
crystallize personal culture to an extent that renders later adult
experiences on political dispositions less influential in explaining
overall differences.

@@

Despite these limitations, we believe our method and findings have
implications for social science research. Sociologists interested in
understanding cultural differences have largely asked about the
\emph{existence} of cultural change in adults (e.g., Kiley and Vaisey
2020; Lersch 2023; Vaisey and Kiley 2021). In any population,
\emph{some} degree of adult cultural change is inevitable. Although
ideal types like the settled dispositions and active updating models are
useful, no theoretical perspective would expect either early life
socialization or adult intrapersonal change to be the sole source of
one's personal culture. Our results reinforce this point, showing the
relevance of both factors and allowing their precise quantification.

This and other recent findings (Quinn et al. 2023, 2024; Stewart and
Berkman 2023) suggest that it is theoretically more productive to
measure the relative importance of these two components in concrete,
substantive cases. Drawing a unified conclusion from survey items
measuring various cultural forms on different scales and across
different time frames is challenging. Nevertheless, the general pattern
suggests that, for most items, intrapersonal change in adulthood is not
the primary reason for the differences we see between people in the
world.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-brett2021}{}}%
Brett, Gordon, and Andrew Miles. 2021. {``Who {Thinks} {How}? {Social}
{Patterns} in {Reliance} on {Automatic} and {Deliberate} {Cognition}.''}
\emph{Sociological Science} 8.

@@

\leavevmode\vadjust pre{\hypertarget{ref-quinn2024}{}}%
Quinn, Joseph M., Robert E. Freeland, E. K. Maloney, Kimberly B. Rogers,
and Lynn Smith-Lovin. 2024. {``Meaning {Change} in {U}.{S}.
{Occupational} {Identities} During the {COVID}-19 {Pandemic}: {Was} {It}
{Temporary} or {Durable}?''} \emph{Social Psychology Quarterly}
01902725241228529. doi:
\href{https://doi.org/10.1177/01902725241228529}{10.1177/01902725241228529}.
